Efficient implementation of penalized regression for genetic risk prediction

### 2018-08-22 paper/draft.tex
save

\title{Efficient implementation of penalized models and their application to genetic risk prediction}
\author{Florian Priv\'e\,$^{\text{ 1,}*}$, Hugues Aschard\,$^{\text{2}}$ and Michael G.B. Blum\,$^{\text{1,}*}$}

@@

Polygenic Risk Scores (PRSs) consist in combining the information across many single-nucleotide polymorphisms (SNPs) in a score reflecting the genetic risk of developing a disease. PRSs are useful for different applications, such as testing the polygenicity of one disease and finding a common genetic contribution between two diseases \cite[]{purcell2009common}. Personalized medicine is another major application of PRSs. Personalized medicine envisions to use PRSs in screening campaigns in order to identify high-risk individuals for a given disease \cite[]{chatterjee2016developing}. As an example of practical application, targeting screening to men at higher polygenic risk could reduce the problem of overdiagnosis and lead to a better benefit-to-harm balance in screening for prostate cancer \cite[]{pashayan2015implications}. 
Yet, PRSs would have to show a high discriminative power between cases and controls in order to be used for helping in the diagnosis of diseases. For example, for screening high-risk individuals and for presymptomatic diagnosis of the general population, it is suggested that the AUC must be greater than 75\% and 99\% respectively \cite[]{janssens2007impact}.

@@

We recently developed two R packages, bigstatsr and bigsnpr, for efficiently analyzing large-scale genome-wide data \cite[]{prive2017efficient}. Package bigstatsr now includes an efficient algorithm with a new implementation for computing sparse linear and logistic regressions on huge datasets as large as the UK Biobank \cite[]{bycroft2017genome}.
In this paper, we present a comprehensive comparative study of our implementation of penalized logistic regression against the C+T method and the T-Trees algorithm, a derivation of random forests that has shown high predictive performance \cite[]{botta2014exploiting}.
In this comparison, we do not include any LMM method for the reasons mentioned before and do not include any SVM method because it is expected to give similar results to logistic regression \cite[]{abraham2012sparsnp}. 
For the C+T model, we report results for a large grid of hyper-parameters. 
For the penalized logistic regression, the choice of hyper-parameters is left to the algorithm so that we report only one model for each simulation. We also use this penalized logistic regression on a feature-augmented dataset in order to capture not only linear effects, but also recessive and dominant effects.

For the simulations of this paper, we use real genotype data and simulate new phenotypes \cite[]{zhou2013polygenic}. 
In order to make our comparison as comprehensive as possible, we compare different disease architectures by varying the number, size and location of causal effects as well as the heritability. We also compare different models for simulating new phenotypes, one with only linear effects, and one that combines linear, dominant and interaction-type effects.
Overall, we found that the penalized logistic regression consistently achieves higher predictive performance than the C+T and T-Trees methods while being very fast. This demonstrates the feasibility and relevance of this approach for PRS computation on large modern datasets.

@@

We use real genotypes of European individuals from a case-control celiac disease study \cite[]{dubois2010multiple}. The composition of this dataset is presented in table \ref{tab:celiac-data}. Details of quality control and imputation for this dataset are available in \cite{prive2017efficient}. For simulations presented later, we first restrict this dataset to controls in order to remove the genetic structure induced by the celiac disease status. Then, we decided to remove population structure because it can affect the predictive performance of methods \cite[]{martin2017human}.
In order to alleviate population structure, we keep people from the UK only and we further remove outliers based on the robust Mahalanobis distances computed using the first 10 principal components of the remaining individuals. This 3-step filtering process results in a sample of 7102 individuals with minimal population structure (see supplementary notebook ``preprocessing''). We also use this dataset for real data application, in this case keeping all 15,155 individuals. Both datasets contain 281,122 SNPs.

@@

We simulate binary phenotypes using a Liability Threshold Model (LTM) with a prevalence of 30\% \cite[]{falconer1965inheritance}. We vary parameters of the simulations in order to match a range of genetic architecture from low to high polygenicity. 
This is achieved by varying the number of causal variants and their location (30, 300, or 3000 anywhere in all 22 chromosomes or 30 in the HLA region of chromosome 6), and the heritability (50\% or 80\%). 
In a second phase, in order to further increase the proportion of causal variants, we restrict the dataset to chromosome 6 only (18,941 SNPs) instead of using all 22 automosal chromosomes (281,122 SNPs).
We also consider deviation from the standard normal additive model, drawing effects of causal SNPs from either a Normal or from a Laplace distribution, and computing liability scores either from a ``simple'' model with linear effects only or a ``fancy'' model that combines linear, dominant and interaction-type effects. For the ``simple'' model, we compute the liability score of the i-th individual $$y_i = \sum_{j\in S_\text{causal}} w_j \cdot \widetilde{G_{i,j}} ~~+~~ \epsilon_i~,$$ where $w_j$ are weights generated from a Gaussian or a Laplace distribution, $G_{i,j}$ is the allele count of individual $i$ for SNP $j$, $\widetilde{G_{i,j}}$ corresponds to its standardized version (zero mean and unit variance for all SNPs) and $\epsilon \sim N(0, 1 - h^2)$.
For the ``fancy'' model, we separate the causal SNPs in three equal sets $S_\text{causal}^{(1)}$, $S_\text{causal}^{(2)}$
and $S_\text{causal}^{(3)}$; $S_\text{causal}^{(3)}$ is further separated in two equal sets, $S_\text{causal}^{(3.1)}$ and $S_\text{causal}^{(3.2)}$.
We then compute $$y_i = \underbrace{\sum_{j\in S_\text{causal}^{(1)}} w_j \cdot \widetilde{G_{i,j}}}_\text{linear} + \underbrace{\sum_{j\in S_\text{causal}^{(2)}} w_j \cdot \widetilde{D_{i,j}}}_\text{dominant} + \underbrace{\sum_{\substack{k=1 \\ j_1=e_k^{(3.1)} \\ j_2=e_k^{(3.2)}}}^{k=\left|S_\text{causal}^{(3.1)}\right|} w_{j_1} \cdot \widetilde{G_{i,j_1} G_{i,j_2}}}_\text{interaction} ~~+~~ \epsilon_i~,$$
where $D_{i,j} = \mathds{1}\left\{G_{i,j} \neq 0\right\}$ and $S_\text{causal}^{(q)}=\left\{e_k^{(q)},~k \in \left\{1, \ldots, \left|S_\text{causal}^{(q)}\right|\right\}\right\}$.
%Note that, because we use the raw genotype counts when coding the interaction terms, the corresponding SNPs also display a marginal effect.

@@

The C+T (Clumping + Thresholding) method directly derives a Polygenic Risk Score (PRS) from the results of Genome-Wide Associations Studies (GWAS). In GWAS, a coefficient of regression (i.e.\ the estimated effect size $\hat\beta_j$) is learned independently for each SNP $j$ along with a corresponding p-value $p_j$. The SNPs are first clumped (C) so that there remain only loci that are weakly correlated with one another (this set of SNPs is denoted $S_\text{clumping}$). Then, thresholding (T) consists in removing SNPs with p-values larger than a threshold $p_T$ to be determined. Finally, a PRS is defined as the sum of allele counts of the remaining SNPs weighted by the corresponding effect coefficients: $$\rm{PRS}_i = \sum_{\substack{j \in S_\text{clumping} \\ p_j~<~p_T}} \hat\beta_j \cdot G_{i,j}~,$$ where $\hat\beta_j$ ($p_j$) are the effect sizes (p-values) learned from the GWAS. In this study, we mostly report scores for a clumping threshold at $r^2 > 0.2$ within regions of 500kb, but we also investigate thresholds of $0.05$ and $0.8$.

@@

Finally, for the penalized logistic regression, we find regression coefficients $\beta_0$ and $\beta$ that minimize the following regularized loss function: $$L(\lambda, \alpha) = \underbrace{ -\sum_{i=1}^n \left( y_i \log\left(p_i\right) + (1 - y_i) \log\left(1 - p_i\right) \right) }_\text{Loss function}   +   \underbrace{ \lambda \left((1-\alpha)\frac{1}{2}\|\beta\|_2^2 + \alpha \|\beta\|_1\right) }_\text{Penalization} ~,$$ where $p_i=1/\left(1+\exp\left(-(\beta_0 + x_i^T\beta)\right)\right)$, $x$ is denoting the genotypes and covariables (e.g.\ principal components), $y$ is the disease status we want to predict, $\lambda$ and $\alpha$ are two regularization parameters that need to be chosen.
Different regularizations can be used to prevent overfitting, among other benefits: the L2-regularization (``ridge'', \cite{hoerl1970ridge}) shrinks coefficients and is ideal if there are many predictors drawn from a Gaussian distribution (corresponds to $\alpha = 0$ in the previous equation);
the L1-regularization (``lasso'', \cite{tibshirani1996regression}) forces some of the coefficients to be equal to zero and can be used as a means of variable selection, leading to sparse models (corresponds to $\alpha = 1$);
the L1- and L2-regularization (``elastic-net'', \cite{zou2005regularization}) is a compromise between the two previous penalties and is particularly useful in the $m \gg n$ situation ($m$: number of SNPs), or any situation involving many correlated predictors (corresponds to $0 < \alpha < 1$) \cite[]{friedman2010regularization}. In this study, we use elastic-net regularization with $\alpha \in \{1, 0.5, 0.05, 0.001\}$ (determined thanks to an embedded grid search).

@@

In order to capture recessive and dominant effects in addition to linear effects, we use feature engineering: we construct a separate dataset with, for each SNP variable, two more variables coding for recessive and dominant effects. This results in a dataset with 3 times as many variables as the initial one, on which we can apply penalized logistic regression. We refer to this method ``logit-triple'' in the results.

@@

When simulating a model with 30 causal SNPs and an heritability of 80\%, penalized logistic regression provides AUC greater than 93\%, nearly reaching the maximum achievable AUC of 94\%, whereas AUC values obtained with C+T method range between 83\% and 90\% (Figure \ref{fig:main-AUC-logit}). 
Moreover, penalized logistic regression consistently provides better predictive performance than the C+T method across all scenarios we considered, except in some case of high polygenicity or small sample size, in which cases all methods perform poorly (AUC values below 60\% -- figures \ref{fig:main-AUC-ntrain} and \ref{fig:supp-AUC-logit}).

@@

In practice, a particular value of the threshold of inclusion of SNPs should be chosen for the C+T method and this choice can dramatically impact the predictive performance of C+T. For example, AUC ranges from less than 60\% to 90\% in a model with only 30 causal SNPs (Figures \ref{fig:main-AUC-PRS} and \ref{fig:supp-AUC-PRS}).

Concerning the $r^2$ threshold of the clumping step in C+T, we mostly used the commonly used value of $0.2$. Yet, using a more stringent value of $0.05$ seems to globally provide higher predictive performance than using $0.2$ (Figures \ref{fig:supp-AUC-all-r2}, \ref{fig:main-AUC-ntrain} and \ref{fig:supp-AUC-chr6-all-r2})

Method ``logit-simple'' that automatically chooses hyper-parameter $\lambda$ provides similar predictive performance than the best predictive performance of the biglasso implementation \cite[]{zeng2017efficient}, only slightly better for biglasso but likely due to over-fitting when reporting the best prediction (Figure \ref{fig:supp-biglasso}) [GARDER? -> SI OUI, A RELANCER].

@@

Joint logistic regressions provide higher AUC values for the Celiac data: 88.8\% with ``logit-simple'' and 89.2\% with ``logit-triple'' as compared to 82.4\% with the C+T method.
The relative increase in partial AUC, for specificities larger than 90\%, is even larger (45\% and 50\%) with partial AUC values of 0.0414, 0.0429 and 0.0286 obtained with ``logit-simple'', ``logit-triple'' and the C+T method, respectively.
Moreover, logistic regressions use less predictors, respectively 3320, 4470 and 9850 (Table \ref{tab:results-celiac}, figures \ref{fig:celiac-roc} and supplementary notebook ``results-celiac'').
Note that for the C+T method, we still report the best result among 102 p-value thresholds.
In terms of computation time, we report only the GWAS computation for the C+T method and we show that the ``logit-simple'' method, while learning jointly on all SNPs at once, is faster than the C+T method (84 vs 148 seconds), and the ``logit-triple'' takes less than twice as long as the ``logit-simple'' (141 vs 84 seconds). [CHANGE: A RELANCER AVEC GRID-SEARCH]

\begin{table}[h]
\caption{Results for the real Celiac dataset. The results are averaged over 100 runs where the training step is randomly composed of 12,000 individuals. In the parentheses is reported the standard deviation of $10^5$ bootstrap samples of the mean of the corresponding variable. Results are reported with 3 significant digits.\label{tab:results-celiac}}
\vspace*{0.5em}
\centering
\begin{tabular}{|l|c|c|c|c|}
  \hline
Method & AUC & pAUC & \# predictors & Execution time (s) \\
  \hline
PRS-max & 0.824 (0.000704) & 0.0286 (0.00016) & 9850 (781) & 148 (0.414) \\
logit-simple & 0.888 (0.000468) & 0.0414 (0.000164) & 3220 (62) & 83.8 (1.27) \\
logit-triple & 0.892 (0.000488) & 0.0429 (0.000174) & 4470 (80.6) & 141 (1.85) \\
   \hline
\end{tabular}
\end{table}

@@

The choice of hyper-parameters' values is very important as it can greatly impact method performance. In the C+T method, there are two main hyper-parameters: the $r^2$ and the $p_T$ thresholds that control how stringent are the clumping and thresholding steps, respectively. 
The choice of the $r^2$ threshold of the clumping step is important. Indeed, on the one hand, choosing a low value for this threshold may discard independently predictive SNPs that are in LD; yet, on the other hand, when choosing a high value for this threshold, too much redundant information would be included in the model, which would bias SNP effects.
Yet, based on the simulations of this paper, it seems that using stringent threshold ($r^2 = 0.05$) leads to better predictive performance, even in the case where causal SNPs are correlated. It means that accurately estimating SNP effects is more important than including all causal SNPs.
Moreover, in this paper, we reported the maximum AUC of 102 $p_T$ thresholds, a threshold that should normally be learned on the training set only. The choice of this threshold is very important as it can greatly impact the predictive performance of the C+T method, which we confirm in this study \cite[]{ware2017heterogeneity}.

On the contrary, in the penalized logistic regression presented here, we developed an automatic procedure called Cross-Model Selection and Averaging (CMSA) that releases investigators from the burden of choosing hyper-parameter $\lambda$ that accounts for the amount of regularization used in the model. Not only this procedure provides near-optimal results, but it also accelerates the training of the model thanks to the development of an early stopping criterion. 
Usually, cross-validation is used to choose hyper-parameter value and then the model is trained again with this particular hyper-parameter value \cite[]{Hastie2008,wei2013large}. Yet, performing cross-validation and retraining is computationally demanding, CMSA offers a less burdensome alternative. 
Concerning hyper-parameter $\alpha$ that accounts for the relative importance of the L1 and L2 regularizations, we use a grid  search directly embedded in the CMSA procedure. 
Our penalized logistic regression implementation does not allow for setting $\alpha = 0$, a model that would assume the infinitesimal model. 
Yet, if using e.g. $\alpha = 0.001$, it would lead to a similar model.

@@

\newpage
\begin{figure}[h]
\centerline{\includegraphics[width=\textwidth]{main-AUC-logit}}
\caption{Main comparison of the C+T method and ``logit-simple'' in scenario \textnumero1. Mean of AUC over 100 simulations for ``logit-simple'' and the maximum AUC reported with the C+T method (``PRS-max''). Upper (lower) panel is presenting results for effets following a Gaussian (Laplace) distribution. Error bars are representing $\pm 2 \text{SD}$ of $10^5$ non-parametric bootstrap of the mean of AUC. The blue dotted line represents the maximum achievable AUC. These results corresponds to the ``simple'' model and an heritability of 80\%.}
\label{fig:main-AUC-logit}
\end{figure}

\newpage
\begin{figure}[h]
\centerline{\includegraphics[width=\textwidth]{main-AUC-PRS}}
\caption{Comparison of three different threshold used in the C+T method in scenario \textnumero1. Mean of AUC over 100 simulations. Upper (lower) panel is presenting results for effets following a Gaussian (Laplace) distribution. Error bars are representing $\pm 2 \text{SD}$ of $10^5$ non-parametric bootstrap of the mean of AUC. The blue dotted line represents the maximum achievable AUC. These results corresponds to the ``simple'' model and an heritability of 80\%.}
\label{fig:main-AUC-PRS}
\end{figure}

\newpage
\begin{figure}[h]
\centerline{\includegraphics[width=\textwidth]{main-AUC-ntrain}}
\caption{Comparison of models when varying sample size in scenario \textnumero3. Mean of AUC over 100 simulations for the maximum values of PRSs for three different $r^2$ thresholds ($0.05$, $0.2$ and $0.8$) and ``logit-simple'' as a function of the training size. These results are for the ``simple'' model with 300 causal SNPs sampled anywhere on the genome. Upper (lower) panels are presenting results for effets following a Gaussian (Laplace) distribution and left (right) panels are presenting results for an heritability of 0.5 (0.8). Error bars are representing $\pm 2 \text{SD}$ of $10^5$ non-parametric bootstrap of the mean of AUC. The blue dotted line represents the maximum achievable AUC.}
\label{fig:main-AUC-ntrain}
\end{figure}

@@

@@

\newpage
\begin{figure}[h]
\centerline{\includegraphics[width=\textwidth]{supp-AUC-logit}}
\caption{Comparison of the C+T method and ``logit-simple'' in scenario \textnumero1. Mean of AUC over 100 simulations for ``logit-simple'' and the maximum AUC reported with the C+T method (``PRS-max''). Upper (lower) panel is presenting results for effets following a Gaussian (Laplace) distribution. Error bars are representing $\pm 2 \text{SD}$ of $10^5$ non-parametric bootstrap of the mean of AUC. The blue dotted line represents the maximum achievable AUC. These results correspond to the ``simple'' model and an heritability of 50\%.}
\label{fig:supp-AUC-logit}
\end{figure}

>>>>

\title{Efficient implementation of penalized regression for genetic risk prediction}
\author{Florian Priv\'e\,$^{\text{ 1,}*}$, Hugues Aschard\,$^{\text{2}}$ and Michael G.B. Blum\,$^{\text{1,}*}$}

@@

Polygenic Risk Scores (PRSs) consist in combining the information across many single-nucleotide polymorphisms (SNPs) in a score reflecting the genetic risk of developing a disease. PRSs are useful for genetic epidemiology when testing the polygenicity of one disease and finding a common genetic contribution between two diseases \cite[]{purcell2009common}. Personalized medicine is another major application of PRSs. Personalized medicine envisions to use PRSs in screening campaigns in order to identify high-risk individuals for a given disease \cite[]{chatterjee2016developing}. As an example of practical application, targeting screening to men at higher polygenic risk could reduce the problem of overdiagnosis and lead to a better benefit-to-harm balance in screening for prostate cancer \cite[]{pashayan2015implications}. 
Yet, PRSs would have to show a high discriminative power between cases and controls in order to be used for helping in the diagnosis of diseases. For example, for screening high-risk individuals and for presymptomatic diagnosis of the general population, it is suggested that the AUC must be greater than 75\% and 99\% respectively \cite[]{janssens2007impact}.

@@

We recently developed two R packages, bigstatsr and bigsnpr, for efficiently analyzing large-scale genome-wide data \cite[]{prive2017efficient}. Package bigstatsr now includes an efficient algorithm with a new implementation for computing sparse linear and logistic regressions on huge datasets as large as the UK Biobank \cite[]{bycroft2017genome}.
In this paper, we present a comprehensive comparative study of our implementation of penalized logistic regression against the C+T method and the T-Trees algorithm, a derivation of random forests that has shown high predictive performance \cite[]{botta2014exploiting}.
In this comparison, we do not include any LMM method for the reasons mentioned before and do not include any SVM method because it is expected to give similar results to logistic regression \cite[]{abraham2012sparsnp}. 
For the C+T model, we report results for a large grid of hyper-parameters. 
For the penalized logistic regression, the choice of hyper-parameters is included in the algorithm so that we report only one model for each simulation. We also use this penalized logistic regression on a feature-augmented dataset in order to capture not only linear effects, but also recessive and dominant effects.

To perform simulations, we use real genotype data and simulate new phenotypes \cite[]{zhou2013polygenic}. 
In order to make our comparison as comprehensive as possible, we compare different disease architectures by varying the number, size and location of causal effects as well as the heritability. We also compare different models for simulating phenotypes, one with only linear effects, and one that combines linear, dominant and interaction-type effects.
Overall, we found that the penalized logistic regression consistently achieves higher predictive performance than the C+T and T-Trees methods while being very fast. This demonstrates the feasibility and relevance of this approach for PRS computation on large modern datasets.

@@

We use real genotypes of European individuals from a case-control celiac disease study \cite[]{dubois2010multiple}. The composition of this dataset is presented in table \ref{tab:celiac-data}. Details of quality control and imputation for this dataset are available in \cite{prive2017efficient}. For simulations presented later, we first restrict this dataset to controls in order to remove the genetic structure induced by the celiac disease status. Then, we decided to remove population structure because it can affect the predictive performance of methods \cite[]{martin2017human}.
In order to alleviate population structure, we keep people from the UK only and we further remove outliers based on a robust Mahalanobis distances computed using the first 10 principal components of the remaining individuals. This 3-step filtering process results in a sample of 7102 individuals with minimal population structure (see supplementary notebook ``preprocessing''). We also use this dataset for real data application, in this case keeping all 15,155 individuals (4496 cases and 10,659 controls). Both datasets contain 281,122 SNPs.

@@

We simulate binary phenotypes using a Liability Threshold Model (LTM) with a prevalence of 30\% \cite[]{falconer1965inheritance}. We vary parameters of the simulations in order to match a range of genetic architecture from low to high polygenicity. 
This is achieved by varying the number of causal variants and their location (30, 300, or 3000 anywhere in all 22 chromosomes or 30 in the HLA region of chromosome 6), and the heritability (50\% or 80\%). 
In a second phase, in order to further increase the proportion of causal variants, we restrict the dataset to chromosome 6 only (18,941 SNPs) instead of using all 22 automosal chromosomes (281,122 SNPs).
We also consider deviation from the standard normal additive model, drawing effects of causal SNPs from either a Normal or from a Laplace distribution, and computing liability scores either from a ``simple'' model with linear effects only or a ``fancy'' model that combines linear, dominant and interaction-type effects. For the ``simple'' model, we compute the liability score of the i-th individual $$y_i = \sum_{j\in S_\text{causal}} w_j \cdot \widetilde{G_{i,j}} ~~+~~ \epsilon_i~,$$ where $w_j$ are weights generated from a Gaussian or a Laplace distribution, $G_{i,j}$ is the allele count of individual $i$ for SNP $j$, $\widetilde{G_{i,j}}$ corresponds to its standardized version (zero mean and unit variance for all SNPs), $\epsilon$ follows a Gaussian distribution $N(0, 1 - h^2)$ and $S_\text{causal}$ is the set of causal SNPs.
For the ``fancy'' model, we simulate phenotypes using linear, dominant and interaction-type effects (see Supplementary Materials).

@@

The C+T (Clumping + Thresholding) method directly derives a Polygenic Risk Score (PRS) from the results of Genome-Wide Associations Studies (GWAS). In GWAS, a coefficient of regression (i.e.\ the estimated effect size $\hat\beta_j$) is learned independently for each SNP $j$ along with a corresponding p-value $p_j$. The SNPs are first clumped (C) so that there remain only loci that are weakly correlated with one another (this set of SNPs is denoted $S_\text{clumping}$). Then, thresholding (T) consists in removing SNPs with p-values larger than a threshold $p_T$ to be determined. Finally, a PRS is defined as the sum of allele counts of the remaining SNPs weighted by the corresponding effect coefficients $$\rm{PRS}_i = \sum_{\substack{j \in S_\text{clumping} \\ p_j~<~p_T}} \hat\beta_j \cdot G_{i,j}~,$$ where $\hat\beta_j$ ($p_j$) are the effect sizes (p-values) learned from the GWAS. In this study, we mostly report scores for a clumping threshold at $r^2 > 0.2$ within regions of 500kb, but we also investigate thresholds of $0.05$ and $0.8$.

@@

Finally, for the penalized logistic regression, we find regression coefficients $\beta_0$ and $\beta$ that minimize the following regularized loss function $$L(\lambda, \alpha) = \underbrace{ -\sum_{i=1}^n \left( y_i \log\left(p_i\right) + (1 - y_i) \log\left(1 - p_i\right) \right) }_\text{Loss function}   +   \underbrace{ \lambda \left((1-\alpha)\frac{1}{2}\|\beta\|_2^2 + \alpha \|\beta\|_1\right) }_\text{Penalization} ~,$$ where $p_i=1/\left(1+\exp\left(-(\beta_0 + x_i^T\beta)\right)\right)$, $x$ is denoting the genotypes and covariables (e.g.\ principal components), $y$ is the disease status to predict, $\lambda$ and $\alpha$ are two regularization parameters that need to be chosen.
Different regularizations can be used to prevent overfitting, among other benefits: the L2-regularization (``ridge'', \cite{hoerl1970ridge}) shrinks coefficients and is ideal if there are many predictors drawn from a Gaussian distribution (corresponds to $\alpha = 0$ in the previous equation);
the L1-regularization (``lasso'', \cite{tibshirani1996regression}) forces some of the coefficients to be equal to zero and can be used as a means of variable selection, leading to sparse models (corresponds to $\alpha = 1$);
the L1- and L2-regularization (``elastic-net'', \cite{zou2005regularization}) is a compromise between the two previous penalties and is particularly useful in the $m \gg n$ situation ($m$: number of SNPs), or any situation involving many correlated predictors (corresponds to $0 < \alpha < 1$) \cite[]{friedman2010regularization}. In this study, we use elastic-net regularization with $\alpha \in \{1, 0.5, 0.05, 0.001\}$ (determined thanks to an embedded grid search).

@@

In order to capture recessive and dominant effects in addition to linear effects, we use feature engineering: we construct a separate dataset with, for each SNP variable, two more variables coding for recessive and dominant effects: one variable is coded 1 if homozygous variant and 0 otherwise, and the other is coded 0 for homozygous referent and 1 otherwise. This results in a dataset with 3 times as many variables as the initial one, on which we can apply penalized logistic regression. We refer to this method ``logit-triple'' in the results.

@@

When simulating a model with 30 causal SNPs and an heritability of 80\%, penalized logistic regression provides AUC greater than 93\%, nearly reaching the maximum achievable AUC of 94\%, whereas AUC values obtained with C+T method range between 83\% and 90\% (Figure \ref{fig:main-AUC-logit}). 
Moreover, penalized logistic regression consistently provides better predictive performance than the C+T method across all scenarios we considered, except in some cases of high polygenicity or small sample size where all methods perform poorly (AUC values below 60\% -- figures \ref{fig:main-AUC-ntrain} and \ref{fig:supp-AUC-logit}).

@@

In practice, a particular value of the threshold of inclusion of SNPs should be chosen for the C+T method and this choice can dramatically impact the predictive performance of C+T. For example, in a model with only 30 causal SNPs, AUC ranges from less than 60\% when using all SNPs passing clumping to 90\% if choosing the optimal p-value threshold (Figures \ref{fig:main-AUC-PRS} and \ref{fig:supp-AUC-PRS}).

Concerning the $r^2$ threshold of the clumping step in C+T, we mostly used the commonly used value of $0.2$. Yet, using a more stringent value of $0.05$ provides higher predictive performance than using $0.2$ in most of the cases tested in this paper (Figures \ref{fig:supp-AUC-all-r2}, \ref{fig:main-AUC-ntrain} and \ref{fig:supp-AUC-chr6-all-r2})

Method ``logit-simple'' that automatically chooses hyper-parameter $\lambda$ provides similar predictive performance than the best predictive performance of the biglasso implementation \cite[]{zeng2017efficient}, only slightly better for biglasso but likely due to over-fitting when reporting the best prediction (Figure \ref{fig:supp-biglasso}).

@@

Joint logistic regressions provide higher AUC values for the Celiac data: 88.7\% with ``logit-simple'' and 89.1\% with ``logit-triple'' as compared to 82.5\% with the C+T method.
The relative increase in partial AUC, for specificities larger than 90\%, is even larger (42\% and 47\%) with partial AUC values of 0.0411, 0.0426 and 0.0289 obtained with ``logit-simple'', ``logit-triple'' and the C+T method, respectively.
Moreover, logistic regressions use less predictors, respectively 1570, 2260 and 8360 (Table \ref{tab:results-celiac}, figures \ref{fig:celiac-roc} and supplementary notebook ``results-celiac'').
Note that for the C+T method, we still report the best result among 102 p-value thresholds.
In terms of computation time, we report only the GWAS computation for the C+T method and we show that the ``logit-simple'' method, while learning jointly on all SNPs at once and testing different hyper-parameter values, is almost as fast as the C+T method (190 vs 130 seconds), and the ``logit-triple'' takes less than twice as long as the ``logit-simple'' (296 vs 190 seconds).

\begin{table}[h]
\caption{Results for the real Celiac dataset. The results are averaged over 100 runs where the training step is randomly composed of 12,000 individuals. In the parentheses is reported the standard deviation of $10^5$ bootstrap samples of the mean of the corresponding variable. Results are reported with 3 significant digits.\label{tab:results-celiac}}
\vspace*{0.5em}
\centering
\begin{tabular}{|l|c|c|c|c|}
  \hline
Method & AUC & pAUC & \# predictors & Execution time (s) \\
  \hline
PRS-max & 0.825 (0.000664) & 0.0289 (0.000187) & 8360 (744) & 130 (0.143) \\ 
logit-simple & 0.887 (0.00061) & 0.0411 (0.000224) & 1570 (46.4) & 190 (1.21) \\ 
logit-triple & 0.891 (0.000628) & 0.0426 (0.000219) & 2260 (56.1) & 296 (2.03) \\ 
   \hline
\end{tabular}
\end{table}

@@

The choice of hyper-parameter values is very important as it can greatly impact method performance. In the C+T method, there are two main hyper-parameters: the $r^2$ and the $p_T$ thresholds that control how stringent are the clumping and thresholding steps, respectively. 
The choice of the $r^2$ threshold of the clumping step is important. Indeed, on the one hand, choosing a low value for this threshold may discard independently predictive SNPs that are in LD; yet, on the other hand, when choosing a high value for this threshold, too much redundant information would be included in the model, which would bias SNP effects.
Based on the simulations, we find that using stringent threshold ($r^2 = 0.05$) leads to better predictive performance, even when causal SNPs are correlated. It means that accurately estimating SNP effects is more important than including all causal SNPs.
Moreover, in this paper, we reported the maximum AUC of 102 different p-value thresholds, a threshold that should normally be learned on the training set only. The choice of this threshold is very important as it can greatly impact the predictive performance of the C+T method, which we confirm in this study \cite[]{ware2017heterogeneity}.

On the contrary, in the penalized logistic regression presented here, we developed an automatic procedure called Cross-Model Selection and Averaging (CMSA) that releases investigators from the burden of choosing hyper-parameter $\lambda$ that accounts for the amount of regularization used in the model. Not only this procedure provides near-optimal results, but it also accelerates the training of the model thanks to the development of an early stopping criterion. 
Usually, cross-validation is used to choose hyper-parameter value and then the model is trained again with this particular hyper-parameter value \cite[]{Hastie2008,wei2013large}. Yet, performing cross-validation and retraining is computationally demanding, CMSA offers a less burdensome alternative. 
Concerning hyper-parameter $\alpha$ that accounts for the relative importance of the L1 and L2 regularizations, we use a grid  search directly embedded in the CMSA procedure. 
%Our penalized logistic regression implementation does not allow for setting $\alpha = 0$, a model that would assume the infinitesimal model. Yet, if using e.g. $\alpha = 0.001$, it would lead to a similar model.

@@

\newpage
\begin{figure}[h]
\centerline{\includegraphics[width=\textwidth]{main-AUC-logit}}
\caption{Main comparison of the C+T and ``logit-simple'' methods in scenario \textnumero1 for the ``simple'' model and an heritability of 80\%. Mean of AUC over 100 simulations for ``logit-simple'' and the maximum AUC reported with the C+T method (``PRS-max''). Upper (lower) panel is presenting results for effets following a Gaussian (Laplace) distribution. Error bars are representing $\pm 2 \text{SD}$ of $10^5$ non-parametric bootstrap of the mean of AUC. The blue dotted line represents the maximum achievable AUC.}
\label{fig:main-AUC-logit}
\end{figure}

\newpage
\begin{figure}[h]
\centerline{\includegraphics[width=\textwidth]{main-AUC-PRS}}
\caption{Comparison of three different p-value thresholds used in the C+T method in scenario \textnumero1 for the ``simple'' model and an heritability of 80\%. Mean of AUC over 100 simulations. Upper (lower) panel is presenting results for effets following a Gaussian (Laplace) distribution. Error bars are representing $\pm 2 \text{SD}$ of $10^5$ non-parametric bootstrap of the mean of AUC. The blue dotted line represents the maximum achievable AUC.}
\label{fig:main-AUC-PRS}
\end{figure}

\newpage
\begin{figure}[h]
\centerline{\includegraphics[width=\textwidth]{main-AUC-ntrain}}
\caption{Comparison of models when varying sample size in scenario \textnumero3 for the ``simple'' model with 300 causal SNPs sampled anywhere on the genome. Mean of AUC over 100 simulations for the maximum values of PRSs for three different $r^2$ thresholds ($0.05$, $0.2$ and $0.8$) and ``logit-simple'' as a function of the training size. Upper (lower) panels are presenting results for effets following a Gaussian (Laplace) distribution and left (right) panels are presenting results for an heritability of 0.5 (0.8). Error bars are representing $\pm 2 \text{SD}$ of $10^5$ non-parametric bootstrap of the mean of AUC. The blue dotted line represents the maximum achievable AUC.}
\label{fig:main-AUC-ntrain}
\end{figure}

@@

\subsection*{``Fancy'' model}

For the ``fancy'' model, we separate the causal SNPs in three equal sets $S_\text{causal}^{(1)}$, $S_\text{causal}^{(2)}$
and $S_\text{causal}^{(3)}$; $S_\text{causal}^{(3)}$ is further separated in two equal sets, $S_\text{causal}^{(3.1)}$ and $S_\text{causal}^{(3.2)}$.
We then compute $$y_i = \underbrace{\sum_{j\in S_\text{causal}^{(1)}} w_j \cdot \widetilde{G_{i,j}}}_\text{linear} + \underbrace{\sum_{j\in S_\text{causal}^{(2)}} w_j \cdot \widetilde{D_{i,j}}}_\text{dominant} + \underbrace{\sum_{\substack{k=1 \\ j_1=e_k^{(3.1)} \\ j_2=e_k^{(3.2)}}}^{k=\left|S_\text{causal}^{(3.1)}\right|} w_{j_1} \cdot \widetilde{G_{i,j_1} G_{i,j_2}}}_\text{interaction} ~~+~~ \epsilon_i~,$$
where $D_{i,j} = \mathds{1}\left\{G_{i,j} \neq 0\right\}$ and $S_\text{causal}^{(q)}=\left\{e_k^{(q)},~k \in \left\{1, \ldots, \left|S_\text{causal}^{(q)}\right|\right\}\right\}$.
%Note that, because we use the raw genotype counts when coding the interaction terms, the corresponding SNPs also display a marginal effect.

@@

\newpage
\begin{figure}[h]
\centerline{\includegraphics[width=\textwidth]{supp-AUC-logit}}
\caption{Comparison of the C+T and ``logit-simple'' methods in scenario \textnumero1 for the ``simple'' model and an heritability of 50\%. Mean of AUC over 100 simulations for ``logit-simple'' and the maximum AUC reported with the C+T method (``PRS-max''). Upper (lower) panel is presenting results for effets following a Gaussian (Laplace) distribution. Error bars are representing $\pm 2 \text{SD}$ of $10^5$ non-parametric bootstrap of the mean of AUC. The blue dotted line represents the maximum achievable AUC.}
\label{fig:supp-AUC-logit}
\end{figure}
